Ditch detection in high-resolution digital elevation models

### 2020-06-27 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds changes to article and rejoinder.

There are other studies where geographical data has been used with machine learning to map ditches. \citet{roelens} used the Random Forests algorithm to detect ditches in Belgium, using LiDAR point clouds. They used several different input variables where neighbouring points of a specific LiDAR point were examined. Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii. We reproduce these neighbouring area variables when building the model in our study. In a pilot study, we compared several different algorithms (Random Forests, XGBoost, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results. Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}. Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, a Random Forests model is used to locate the ditches.

We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches. To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment},  where a high precision manual mapping has been performed. This mapping was used to both train the Random Forests model and to ground-truth the models. To evaluate the performance of the model, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.

@@

\subsection{Digitising the ground truth}
A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{Appendix} \ref{lidartodem} for specifications of the process. To train and evaluate the Random Forests model, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised. Road ditches, forest ditches, and agricultural ditches can be found in the catchment. A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches. From this layer, most of the ditches could be easily identified. To ensure that the digitised ditch was located correctly inside the ditch and not on the bank nearby, the digitisation was conducted by zooming in to a scale of 1:600. In addition to this, both current and historical records and orthophotos dating back to 1960’s were used to help locate all the ditches. In uncertain areas, we also found it useful to view the data in 3D. However, the digitising was conducted in 2D, and in uncertain areas the ditches were verified in the field. We are confident that we have captured all the ditches in the catchment according to the digitised  ditch layers in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}. The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold. Though, the length of the stream network vary substantially between season, ranging from 140 km during baseflow to 630 km during high-flow conditions \citep{mappingtemporal}. The digitisation of the ditch networks showed that there were 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}). The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which aims to show median flow conditions, was 197 km.

\begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections. The 10 subsections with a white border were used for developing the model before the experiment. The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation. Each rectangular subsection represents $2997 \cdot 2620$ pixels (roughly 196 hectare).}
    \label{fig:swedenkrycklan}
\end{figure}

From the ditch mapping, the vector layer was rasterised so it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM. Because the observed average width of ditches is larger than 0.5 metres, all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels. \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening. The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests model. The ditches all have a generalised width based on a perceived average of ditch width in the area. Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch. However, this made for a good compromise for the average ditch.

Because our priority was to detect ditches, rather than accurately detecting each pixel labelled as a ditch, some adjustments were made when producing the prediction results. The dataset was divided into a lower resolution grid of $6*6$ pixels (9 $m^2$) for each grid. Each grid cell that contained at least 25\% ditch pixels was then labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}. This layer was used to ground-truth the Random Forests model.

\begin{figure} [htb!]
    \centering
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as input when training the model. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the model.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

@@

\subsubsection{Processing the digital terrain indices}
Developing the Random Forests model involved examining how different kinds of input variables affected the prediction. Several possible data manipulation methods could theoretically produce a better prediction. To improve the prediction, steps were taken where neighbouring pixels were included to give a representation of the area surrounding a specific pixel, similar to the approach of \citet{roelens}.

The digital terrain indices provided a satisfactory foundation for the model, but lacked in the generalisability of their predictions. More diverse input variables were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation. Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis. These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{h}, and \hyperref[fig:features]{j}.

@@

From the Slope and Sky View Factor indices, input variables for the model were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}.

@@

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_j_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_k_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_l_lo.jpg}}}\hspace{5pt}
    \caption{Example of 11 of the 81 input variables, in addition to ditch labels used by the model for a small sample area. The radii represents pixels with a 0.5 m resolution. \newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF Gabor, \textbf{e:} HPMF ditch amplification, \textbf{f:} HPMF ditch amplification - streams removed \textbf{g:} Sky View Factor Gabor, \textbf{h:} Sky View Factor max, radius 6, \textbf{i:} Sky View Factor non-ditch amplification  \textbf{j:} Impoundment mean, radius 3, \textbf{k:} Impoundment ditch amplification, \textbf{l:} Impoundment ditch amplification - streams removed}
    \label{fig:features}
\end{figure}
\clearpage

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our model, the raster and ditch label data of Krycklan were manually divided into 21 subsections.  Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the model to test different ways of preparing the data. This allowed the model to be evaluated on unseen data to strengthen the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.

Using the 11 subsections in the hold-out data for the final experiment, a process called \textit{k-fold cross validation} was employed (11-fold cross validation in our case). K-fold cross validation is a method where you divide your dataset into folds of similar size and train a model on all but one of your folds (subsections). You then use that subsection to evaluate the results \citep{crossvalidation}. Using this technique, shifting which subsection to leave out from the training, allowed us to train 11 different Random Forests classifying models with a large amount of data from the remaining 10 subsections in the hold-out data, producing 11 sub-experiments to evaluate the method on.

\subsection{Building the Random Forests model}

@@

Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}. Subspace sampling is the process of drawing a random subset of variables from the input variable set \citet{ho}. The bagging technique takes, with replacement, a set of different random samples for constructing each tree \citep{flach}. There are several hyperparameters that can be tuned to achieve better results from the model. For example, you can adjust the amount of decision trees to use, or the maximum amount of input variables to use for each subspace sampling \citep{scikit-learn}. We performed a simple hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction. Evaluating a maximum of 25 input variables for each node, and using 200 trees produced the best results. Setting the class weight to \textit{balanced} also improved the performance of the classifier.

@@

The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifier was trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}. The probability predictions used in this Random Forests implementation are computed by counting the number of occurrences of a class output and dividing with the total amount of trees in the forest \citep{scikit-learn}. The testing phase showed that the classifier produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high. A high ratio led to the model not being punished for mislabelling ditches as non-ditches, causing it to prioritise a high accuracy over a high recall. According to \citet{balanced}, an imbalanced training dataset causes a minority class to be less accurately predicted. Because the ditch class is much less common than the non-ditch class, we attempted to train the model with a roughly equal amount of ditch pixels and non-ditch pixels to balance the model. The model was, however, still evaluated on all the pixels in the 11 validation subsections.

\begin{table} [!htb]
    \tbl{The 81 input variables used when training the model.}
    {\begin{tabular}{ll} \toprule
      Variable/Algorithm\textsuperscript{a} & Circular radii\textsuperscript{b} \\ \midrule

@@

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_balanced_masks_A_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the model. \textbf{a, b: }Examples of two subsection from the training dataset.}
    \label{fig:balancedmasks}
\end{figure}

\subsection{Post-processing}
The model outputs a ditch class probability prediction for each pixel. Values close to zero indicate a low probability of a pixel lying in a ditch, whereas values close to one equals a high ditch probability (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{a}).

@@

The third step involved taking measures to try to fill gaps in ditches that the model failed to correctly predict. This was done by aggregating pixels covered by cone masks expanding outwards in different directions from the examined pixel. This step also amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important to help make the next step more effective (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).

@@

\subsubsection{Binarisation with grid zones}
The model's ditch prediction is given in the original resolution raster format. To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data seen in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}. A mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 35 \%. This helped to fill in gaps where lone pixels in ditches had been incorrectly classified, and also helped in the next step of the post-processing (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).

@@

From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the model generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.

\begin{table}[!htb]
    \tbl{Metrics for the prediction performance of our model.}{\begin{tabular}{lccc} 
        \toprule
        Metric & Total\textsuperscript{a} & Subsection\textsuperscript{b}& CI 95\%\textsuperscript{c} \\ 
        &&Average&\\ \midrule
        Accuracy     \% & 98.96 & 98.96 & [98.64 , 99.27] \\
        Recall       \% & 77.41 & 76.81 & [69.35 , 84.27] \\
        Precision    \% & 73.89 & 73.06 & [69.73 , 76.40] \\
        $\kappa$ rating & 0.751 & 0.739 & [0.693 , 0.786] \\
        AUPRC           & 0.759 & 0.752 & [0.706 , 0.798] \\ 
        \bottomrule
    \end{tabular}}
    \tabnote{
        \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
        \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
        \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
    \label{predictionperformance}
\end{table}

@@

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_1_lo.jpg}}}\hspace{7pt}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_2_lo.jpg}}}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_3_lo.jpg}}}\hspace{7pt}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_4_lo.jpg}}}
    \caption{\textbf{a} and \textbf{b} illustrate that despite a point density of 20 points $m^2$, not all points reach the ground in places with dense vegetation. In \textbf{a} there is a clear difference between the smooth surface of the DEM in the open area (black arrow), and in the areas covered by trees and shrubs (red arrows), where the DEM is much coarser due to interpolation with a lower number of LiDAR points here. The orthophoto (\textbf{b}), clearly shows that trees and shrubs are growing in the ditches in this area. This makes it more difficult to detect ditches. However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the model still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
    \label{fig:resultstreesbushes}
\end{figure}

>>>>

There are other studies where geographical data has been used with machine learning to map ditches. \citet{roelens} used the Random Forests algorithm to detect ditches in Belgium, using LiDAR point clouds. They used several different input variables where neighbouring points of a specific LiDAR point were examined. Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii. We reproduce these neighbouring area variables when building the ditch locator in our study. In a pilot study, we compared several different algorithms (Random Forests, XGBoost, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results. Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}. Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, Random Forests models are used to locate the ditches.

We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches. To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed. This mapping was used to both train the Random Forests models and to produce the ground-truth. To evaluate the performance of the ditch locator, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.

@@

\subsection{Digitising the ground truth}
A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{Appendix} \ref{lidartodem} for specifications of the process. To train and evaluate the Random Forests models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised. Road ditches, forest ditches, and agricultural ditches can be found in the catchment. A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches. From this layer, most of the ditches could be easily identified. To ensure that the digitised ditch was located correctly inside the ditch and not on the bank nearby, the digitisation was conducted by zooming in to a scale of 1:600. In addition to this, both current and historical records and orthophotos dating back to 1960’s were used to help locate all the ditches. In uncertain areas, we also found it useful to view the data in 3D. However, the digitising was conducted in 2D, and in uncertain areas the ditches were verified in the field. We are confident that we have captured all the ditches in the catchment according to the digitised  ditch layers in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}. The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold. Though, the length of the stream network vary substantially between season, ranging from 140 km during baseflow to 630 km during high-flow conditions \citep{mappingtemporal}. The digitisation of the ditch networks showed that there were 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}). The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which aims to show median flow conditions, was 197 km.

\begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections. The 10 subsections with a white border were used for developing input variables and optimising parameters before the experiment. The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation. Each rectangular subsection represents $2997 \cdot 2620$ pixels (roughly 196 hectare).}
    \label{fig:swedenkrycklan}
\end{figure}

From the ditch mapping, the vector layer was rasterised so it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM. Because the observed average width of ditches is larger than 0.5 metres, all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels. \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening. The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests models. The ditches all have a generalised width based on a perceived average of ditch width in the area. Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch. However, this made for a good compromise for the average ditch.

Because our priority was to detect ditches, rather than accurately detecting each pixel labelled as a ditch, some adjustments were made when producing the prediction results. The dataset was divided into a lower resolution grid of $6*6$ pixels (9 $m^2$) for each grid. Each grid cell that contained at least 25\% ditch pixels was then labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}. This layer was used to ground-truth the Random Forests models.

\begin{figure} [htb!]
    \centering
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as input when training the models. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the ditch locator.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

@@

\subsubsection{Processing the digital terrain indices}
To improve the performance of the Random Forests models, we examined how different kinds of input variables affected the prediction. Several possible data manipulation methods could theoretically produce a better prediction. To improve the prediction, steps were taken where neighbouring pixels were included to give a representation of the area surrounding a specific pixel, similar to the approach of \citet{roelens}.

The digital terrain indices provided a satisfactory foundation for the models, but lacked in the generalisability of their predictions. More diverse input variables were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation. Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis. These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{h}, and \hyperref[fig:features]{j}.

@@

From the Slope and Sky View Factor indices, input variables for the models were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}.

@@

Overall, 81 input variables were extracted from the terrain indices. We then conducted a sub-experiment to find the best input variables, as well as the optimal number of variables to use. Random Forest's feature importance was used to select the most valuable input variables for different variable amounts. It was found that using 53 input variables produced the best results.

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_j_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_k_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_l_lo.jpg}}}\hspace{5pt}
    \caption{Example of 11 of the 81 input variables, in addition to ditch labels used by the models for a small sample area. The radii represents pixels with a 0.5 m resolution. \newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF Gabor, \textbf{e:} HPMF ditch amplification, \textbf{f:} HPMF ditch amplification - streams removed \textbf{g:} Sky View Factor Gabor, \textbf{h:} Sky View Factor max, radius 6, \textbf{i:} Sky View Factor non-ditch amplification  \textbf{j:} Impoundment mean, radius 3, \textbf{k:} Impoundment ditch amplification, \textbf{l:} Impoundment ditch amplification - streams removed}
    \label{fig:features}
\end{figure}
\clearpage

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our ditch locator, the raster and ditch label data of Krycklan were manually divided into 21 subsections. Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the ditch locator to test different ways of preparing the data. This allowed the ditch locator to be evaluated on unseen data to strengthen the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.

Using the 11 subsections in the hold-out data for the final experiment, a process called \textit{k-fold cross validation} was employed (11-fold cross validation in our case). K-fold cross validation is a method where you divide your dataset into folds of similar size and train a model on all but one of your folds (subsections). You then use that subsection to evaluate the results \citep{crossvalidation}. A new model is trained using the training folds for each iteration (\hyperref[fig:crossvalidation]{Figure} \ref{fig:crossvalidation}). Using this technique, shifting which subsection to leave out from the training, allowed us to train 11 different Random Forests classifying models with a large amount of data from the remaining 10 subsections in the hold-out data, producing 11 independent sub-experiments to evaluate the method on.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\linewidth]{images/cross_validation_lo.jpg}
    \caption{11-fold cross validation. Each subsection in the hold-out data was used once for testing, with the remaining subsections being used to build a new model for each iteration.}
    \label{fig:crossvalidation}
\end{figure}

\subsection{Developing the Random Forests model}

@@

Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}. Subspace sampling is the process of drawing a random subset of variables from the input variable set \citet{ho}. The bagging technique takes, with replacement, a set of different random samples for constructing each tree \citep{flach}. There are several hyperparameters that can be tuned to achieve better results from the model. For example, you can adjust the amount of decision trees to use, or the maximum amount of input variables to use for each subspace sampling \citep{scikit-learn}. We performed a simple hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction. Evaluating a maximum of 25 input variables for each node, and using 200 trees produced the best results. Setting the class weight to \textit{balanced} also improved the performance of the classifiers.

@@

The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}. The probability predictions used in this Random Forests implementation are computed by counting the number of occurrences of a class output and dividing with the total amount of trees in the forest \citep{scikit-learn}. The testing phase showed that the classifiers produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high. A high ratio led to the models not being punished for mislabelling ditches as non-ditches, causing it to prioritise a high accuracy over a high recall. According to \citet{balanced}, an imbalanced training dataset causes a minority class to be less accurately predicted. Because the ditch class is much less common than the non-ditch class, we attempted to train the models with a roughly equal amount of ditch pixels and non-ditch pixels to balance the models. The models were, however, still evaluated on all the pixels in the 11 validation subsections.

\begin{table} [!htb]
    \tbl{The 81 input variables used when training the models.}
    {\begin{tabular}{ll} \toprule
      Variable/Algorithm\textsuperscript{a} & Circular radii\textsuperscript{b} \\ \midrule

@@

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_balanced_masks_A_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the models. \textbf{a, b: }Examples of two subsection from the training dataset.}
    \label{fig:balancedmasks}
\end{figure}

\subsection{Post-processing}
The models output a ditch class probability prediction for each pixel. Values close to zero indicate a low probability of a pixel lying in a ditch, whereas values close to one equals a high ditch probability (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{a}).

@@

The third step involved taking measures to try to fill gaps in ditches that the models failed to correctly predict. This was done by aggregating pixels covered by cone masks expanding outwards in different directions from the examined pixel. This step also amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important to help make the next step more effective (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).

@@

\subsubsection{Binarisation with grid zones}
The models' ditch prediction is given in the original resolution raster format. To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data seen in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}. A mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 35 \%. This helped to fill in gaps where lone pixels in ditches had been incorrectly classified, and also helped in the next step of the post-processing (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).

@@

From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the ditch locator generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.

\begin{table}[!htb]
    \tbl{Metrics for the prediction performance of our ditch locator.}{\begin{tabular}{lccc} 
        \toprule
        Metric & Total\textsuperscript{a} & Subsection\textsuperscript{b}& CI 95\%\textsuperscript{c} \\ 
        &&Average&\\ \midrule
        Accuracy     \% & 98.96 & 98.96 & [98.64 , 99.27] \\
        Recall       \% & 77.41 & 76.81 & [69.35 , 84.27] \\
        Precision    \% & 73.89 & 73.06 & [69.73 , 76.40] \\
        $\kappa$ rating & 0.751 & 0.739 & [0.693 , 0.786] \\
        AUPRC           & 0.759 & 0.752 & [0.706 , 0.798] \\ 
        \bottomrule
    \end{tabular}}
    \tabnote{
        \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
        \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
        \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
    \label{predictionperformance}
\end{table}

@@

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_1_lo.jpg}}}\hspace{7pt}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_2_lo.jpg}}}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_3_lo.jpg}}}\hspace{7pt}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_4_lo.jpg}}}
    \caption{\textbf{a} and \textbf{b} illustrate that despite a point density of 20 points $m^2$, not all points reach the ground in places with dense vegetation. In \textbf{a} there is a clear difference between the smooth surface of the DEM in the open area (black arrow), and in the areas covered by trees and shrubs (red arrows), where the DEM is much coarser due to interpolation with a lower number of LiDAR points here. The orthophoto (\textbf{b}), clearly shows that trees and shrubs are growing in the ditches in this area. This makes it more difficult to detect ditches. However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the ditch locator still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
    \label{fig:resultstreesbushes}
\end{figure}
